---
title: "A Complete Example"
description: "A package from start to finish: Making a heckin' case converter."
weight: 2
---

```{r}
#| label: setup
#| include: false
library(heck)
```

The Rust crate ecosystem is rich with very small and very powerful utility
libraries. One of the most downloaded crates is [heck](https://docs.rs/heck). It
provides traits and structs to perform some of the most common case conversions. 
In this tutorial we'll create a zero-dependency R package to provide the common
case conversions. This is mainly designed to give you a sense of the development
workflow and also showcase what can be done with extendr. The resulting R 
package will be more performant but less flexible than the  R package 
[`{snakecase}`](https://tazinho.github.io/snakecase/).

## Getting started

The first thing we need to do is create a new R package:

```{r}
#| label: create-package
#| eval: false
usethis::create_package("heck")
```

Once we have the basic R package setup, then we need to add the necessary 
`extendr` scaffolding that will link our Rust code to R and ensure the package
builds properly.

```{r}
#| label: use-extendr
#| eval: false
rextendr::use_extendr(crate_name = "rheck", lib_name = "rheck")
```

{% <callout> %}
Normally, the crate and library names will default to the R package name, but in
this case, we cannot do that because it would create a recursive dependency in
the Rust library. To get around this, we name the internal Rust crate and
library `rheck`, which allows us to name the R package `{heck}`.
{% </callout> %}

Next, `heck` is needed as a dependency. The easiest way to do that is to call

```{r}
#| label: use-crate
#| eval: false
rextendr::use_crate("heck")
```

If you prefer to use your terminal, however, simply navigate to `src/rust` and 
run `cargo add heck`. With this, you have everything you need to get started.

## Scalar snek case conversion

```{.rust}
use heck::ToSnekCase;
```

Let's start by creating a simple function to take a single string, and convert
it to snake case. First, the trait `ToSnekCase` needs to be imported so that the
method `to_snek_case()` is available to `&str`.

```{.rust}
use heck::ToSnekCase;

#[extendr]
fn to_snek_case(x: &str) -> String {
    x.to_snek_case()
}
```

Simple enough, right? Let's give it a shot. To make it accessible from your R
session, it needs to be included in your `extendr_module! {}` macro.

``` rust
extendr_module! {
    mod heck;
    fn to_snek_case;
}
```

From your R session, run `devtools::document()` followed by 
`devtools::load_all()` to make the function available. 

```{r}
to_snek_case("MakeMe-Snake case")
```

Rarely is it useful to run a function on just a scalar character value. Rust,
though, works with scalars by default and adding vectorization is another step.

```{r, error = TRUE}
to_snek_case(c("DontStep", "on-Snek"))
```

Providing a character vector causes an error. So how do you go about
vectorizing?

## Vectorizing snek case conversion

To vectorize this function, you need to apply the conversion to each element in 
a character vector. The extendr wrapper struct for a character vector is called 
`Strings`. To take in a character vector and also return one, the function 
signature should look like this:

``` rust
#[extendr]
fn to_snek_case(x: Strings) -> Strings {
}
```

This says there is an argument `x` which must be a character vector and this
function must also return the `Strings` (a character vector). The return type is
signaled by `->`. To iterate through this you can use the `.into_iter()` method 
on the character vector.

``` rust
#[extendr]
fn to_snek_case(x: Strings) -> Strings {
    x.into_iter()
    // the rest of the function
}
```

Iterators have a method called `.map()` (yes, just like `purrr::map()`). It lets
you apply a closure (an anonymous function) to each element of the iterator. In
this case, each element is an
[`Rstr`](https://extendr.github.io/extendr/extendr_api/wrapper/rstr/struct.Rstr.html).
The `Rstr` has a method `.as_str()` which will return a string slice `&str`. You
can take this slice and pass it on to `.to_snek_case()`. After having mapped
over each element, the results are `.collect()`ed into another `Strings`.

```{.rust}
#[extendr]
fn to_snek_case(x: Strings) -> Strings {
    x.into_iter()
        .map(|xi| {
            xi.as_str().to_snek_case()
        })
        .collect::<Strings>()
}
```

This new version of the function can be used in a vectorized manner:

```{r}
to_snek_case(c("DontStep", "on-Snek"))
```

But can it handle a missing value out of the box?

```{r}
to_snek_case(c("DontStep", NA_character_, "on-Snek"))
```

Well, sort of. The `as_str()` method when used on a missing value will return
`"NA"` which is not in a user's best interest.

## Handling missing values

Instead of returning `"na"`, it would be better to return an *actual* missing
value. Those can be created using each scalar's `na()` method e.g. `Rstr::na()`.

You can modify the `.map()` statement to check if an `NA` is present, and, if
so, return an `NA` value. To perform this check, use the `is_na()` method which
returns a `bool`, a value that is either `true` or `false`. The result can then 
be [`match`ed](https://doc.rust-lang.org/book/ch06-02-match.html). When it is
missing, the match arm returns the `NA` scalar value. When it is not missing,
the `Rstr` is converted to snek case. However, since the `true` arm is an `Rstr`
the other `false` arm must *also* be an `Rstr`. To accomplish this use the
`Rstr::from()` method.

```{.rust}
#[extendr]
fn to_snek_case(x: Strings) -> Strings {
    x.into_iter()
        .map(|xi| match xi.is_na() {
            true => Rstr::na(),
            false => Rstr::from(xi.as_str().to_snek_case()),
        })
        .collect::<Strings>()
}
```

This function can now handle missing values!

```{r}
to_snek_case(c("DontStep", NA_character_, "on-Snek"))
```

## Automating code-writing with a macro!

There are traits for the other case conversions such as `ToKebabCase`,
`ToPascalCase`, `ToShoutyKebabCase` and others, with each having a similar 
method name: `.to_kebab_case()`, `to_pascal_case()`, `.to_shouty_kebab_case()`. 
You can either copy the above code for `to_snek_case()` and change the method 
call multiple times, *or* you can use a macro as a form of code generation. A 
macro allows you to generate code in a short hand manner. This macro takes an
identifier which has a placeholder called `$fn_name`: `$fn_name:ident`.

``` rust
macro_rules! make_heck_fn {
    ($fn_name:ident) => {
        /// @export
        #[extendr]
        fn $fn_name(x: Strings) -> Strings {
            x.into_iter()
                .map(|xi| match xi.is_na() {
                    true => Rstr::na(),
                    false => Rstr::from(xi.as_str().$fn_name()),
                })
                .collect::<Strings>()
        }
    };
}
```

The `$fn_name` placeholder is put as the function name definition which is the
same as the method name. To use this macro to generate the rest of the functions
the other traits need to be imported.

```{.rust}
use heck::{
    ToKebabCase, ToShoutyKebabCase,
    ToSnekCase, ToShoutySnakeCase,
    ToPascalCase, ToUpperCamelCase,
    ToTrainCase, ToTitleCase,
};
```

With the traits in scope, the macro can be invoked to generate the other
functions.

``` rust
make_heck_fn!(to_snek_case);
make_heck_fn!(to_shouty_snake_case);
make_heck_fn!(to_kebab_case);
make_heck_fn!(to_shouty_kebab_case);
make_heck_fn!(to_pascal_case);
make_heck_fn!(to_upper_camel_case);
make_heck_fn!(to_train_case);
make_heck_fn!(to_title_case);
```

Note that each of these functions should be added to the `extendr_module! {}`
macro in order for them to be available from R.

Test it out with the `to_shouty_kebab_case()` function!

```{r}
to_shouty_kebab_case("lorem:IpsumDolor__sit^amet")
```

And with that, you've created an R package that provides case conversion using
heck, and you've done it with just a few lines of code!

## Bench marking with `{snakecase}`

To illustrate the performance gains from using a vectorized Rust funciton, a
`bench::mark()` is created between `to_snek_case()` and
`snakecase::to_snake_case()`.

The bench mark will use 5000 randomly generated lorem ipsum sentences.

```{r, warning = FALSE}
set.seed(1701)
x <- unlist(lorem::ipsum(5000, 1, 10))

head(x)

bench::mark(
  rust = to_snek_case(x),
  snakecase = snakecase::to_snake_case(x),
  relative = TRUE
)
```

{% <callout> %}
The memory usage for rust-based functions is likely quite understated. This is
because memory allocated outside of R cannot be tracked.
See [bench::mark](https://bench.r-lib.org/reference/mark.html#value)
documentation for more information.
{% </callout> %}

## The whole thing

In just 42 lines of code (empty lines included), you can create a very
performant R package!

``` rust
use extendr_api::prelude::*;

use heck::{
    ToKebabCase, ToPascalCase, ToShoutyKebabCase, ToShoutySnakeCase, ToSnekCase, ToTitleCase,
    ToTrainCase, ToUpperCamelCase,
};

macro_rules! make_heck_fn {
    ($fn_name:ident) => {
        #[extendr]
        /// @export
        fn $fn_name(x: Strings) -> Strings {
            x.into_iter()
                .map(|xi| match xi.is_na() {
                    true => Rstr::na(),
                    false => Rstr::from(xi.as_str().$fn_name()),
                })
                .collect::<Strings>()
        }
    };
}

make_heck_fn!(to_snek_case);
make_heck_fn!(to_shouty_snake_case);
make_heck_fn!(to_kebab_case);
make_heck_fn!(to_shouty_kebab_case);
make_heck_fn!(to_pascal_case);
make_heck_fn!(to_upper_camel_case);
make_heck_fn!(to_train_case);
make_heck_fn!(to_title_case);

extendr_module! {
    mod heck;
    fn to_snek_case;
    fn to_shouty_snake_case;
    fn to_kebab_case;
    fn to_shouty_kebab_case;
    fn to_pascal_case;
    fn to_upper_camel_case;
    fn to_title_case;
    fn to_train_case;
}
```