---
title: "Introduction"
description: "Write Rust code and call it in R!"
weight: 2
---

```{r}
#| label: create-package
#| include: false
pkgpath <- file.path(tempdir(), "helloextendr")
usethis::create_package(pkgpath, open = FALSE)
```

```{r}
#| label: set-root-dir
#| include: false
knitr::opts_knit$set(root.dir = pkgpath)
```

```{r}
#| label: use-extendr
#| include: false
rextendr::use_extendr()

# need to use dev extendr
rextendr::use_crate("extendr-api", git = "https://github.com/extendr/extendr")
```

```{r}
#| label: document
#| include: false
devtools::document()
devtools::load_all()
```

This is the website for `extendr`, a developer-first utility for building R
packages that call Rust code. With extendr, you can write Rust code like this:

```{r}
#| label: read-lib.rs
#| echo: false
#| attr-output: 'filename="lib.rs"'
#| class-output: rust
libpath <- file.path("src", "rust", "src", "lib.rs")
cat(readLines(libpath), sep = "\n")
```

and then in R you can call it like this:

```{r}
#| label: hello
hello("world")
```

We call the whole project extendr, of course, but it isn't just one thing. It's 
actually a *suite* of software tools. On the Rust side, extendr consists of 
several crates, but you will typically interact with `extender-api`, a 
developer-facing crate providing a host of data structures, traits, and macros — 
like `#[extendr]` in the above example — that make it easy to write Rust code to 
call in R. On the R side, there is `rextendr`, an R package that works sort of 
like `usethis`, sort of like `devtools`, automating many of the repetitive tasks 
required to setup or scaffold your project, allowing you to focus on the 
important business of writing Rust and R code. 

```{=html}
<div class="w-fit mx-auto my-4 md:my-8 flex flex-row items-center gap-2">
  <a href="https://extendr.github.io/extendr/extendr_api/" class="bc btn-lg-outline h-auto p-2 md:p-4 font-mono md:text-[1.5em]">
    <span class="iconify mdi--language-rust"></span>
    extendr-api
  </a>
  <p class="text-gray-600 dark:text-gray-300 m-0 md:text-[1.5em]">&harr;</p>
  <a href="https://extendr.github.io/rextendr/" class="bc btn-lg-outline h-auto p-2 md:p-4 font-mono md:text-[1.5em]">
    rextendr
    <span class="iconify mdi--language-r"></span>
  </a>
</div>
```

So, when you go to setup a repository to develop a Rust-powered R package, you 
will reach for rextendr first. Then, when you start writing Rust code for your R 
package, you will turn to extendr-api.

## Create an R package

The process of building a Rust-powered R package with extendr boils down to 
four simple steps:

1. Setup R package with `usethis::create_package()`.
2. Add extendr scaffolding with `rextendr::use_extendr()`.
3. Write Rust code and export it with `extendr-api`.
4. Build and document with `devtools::document()`.

Once you have worked through those steps, you can then load your package with 
`devtools::load_all()` and start using your package! Here is an example that 
loads the above `lib.rs` into your package at `src/rust/src/lib.rs`, compiles 
it, generates wrappers and documentation, and then calls the `hello()` function 
from R:

```r
# setup R package
usethis::create_package("helloextendr")

# add extendr scaffolding
rextendr::use_extendr()

# build and document package
devtools::document()

# load package
devtools::load_all()

# call rust function
hello("world")
```

```{r}
#| label: hello-again
#| echo: false
hello("world")
```

A Rust-powered R package in one fell swoop! Now, if you want to see a complete 
example of an actual R package, check out the [heck example](@/example.md).

## What's next?

-   If you do not already have Rust, R, and rextendr, check out our [Installation guide](@/installation.md).
-   Want to learn more about Rust? Check out our [Rust tutorials](@/rust-intro/_index.md).
-   Ready to dive in to package development? You will want to consult our
    [Development guide](@/development/_index.md).
-   Have a Rust-powered package and want to publish it to CRAN? Jump to our 
    [Publishing recommendations](@/publishing/_index.md).
-   Here to join the extendr team and contribute to new features and 
    maintenance? Have a look at our [Guide for contributors](@/contributing/_index.md).